Predicting Success of Motivational Interviews with Recurrent Neural Network and Probabilistic Model for Patient-Provider Communication Sequences

### 2017-09-01 paper-amia-summit-2018/amia_summit_2018.tex
update counselor and adolescent dialogue

As can be seen from Table~\ref{tab:data_dist}, our data set is imbalanced and predominately composed of successful sequences with only a small percentage (17.80\%) of unsuccessful sequences. Usually, predictive accuracy misrepresents the performance of an employed algorithm for an imbalanced data. Because simply predict the majority class (successful) would provide a predictive accuracy 82.20\%. Therefore, it is important to handle imbalance data properly. Many solutions are proposed to deal with the imbalanced data sets, which can be divided into two categories: data and algorithmic levels. In this study, we utilized under and over sampling methods at data processing levels for balancing the adolescent obesity data set. Synthetic Minority Over-sampling (SMOTE) method is the most widely used method for over-sampling an imbalanced data set, in which new synthetic examples were generated from minority class \cite{chawla2002smote}. In this study, we generated synthetic examples along the borderline between minority examples and their selected nearest neighbors \cite{nguyen2011borderline}. On the other hand, under sampling method under samples the majority class by replacing a cluster of majority samples by the cluster centroid of a K-Means algorithm. Annotation column in Table~\ref{tab:anno_examp} shows the sequence of behavior codes from top to bottom, where counselor starts with an open-ended question and gets positive feedback at the end. \\

\begin{table}[h]
\caption{Fragment of the annotated transcript of a dialogue between a counselor and an adolescent.}    
\label{tab:anno_examp}
\centering
\begin{tabular}{|l|p{3.6cm}|l|p{8cm}|}
\hline
Annotation  & Description & Speaker & Text \\\hline
331 &	Open-ended question, elicit change talk positive &	Counselor &	do you feel like making healthier choices for your snacks and your meals is something you would be able to do? mm-hmm meaning is that food available for you? \\\hline
117 &	Low Uptake, positive	& Adolescent &	Yes \\\hline
301 &	Structure Session	& Counselor &	okay and thats an important thing for us to think about cause i would not want to help you come up with a plan that you would not be able to do without somebody else help so the last part of your plan is how somebody could be supportive to you meaning how they can help you be successful and so we should choose somebody who you feel like is around often enough \\\hline
112 &	Change Talk positive	& Adolescent &	my um aunt \\\hline
301 &	Structure Session	& Counselor &	okay so lets stick something my aunt can do \\\hline
112 &	Change Talk positive &	Adolescent &	she could when i am doing when i am eating something that i should i could not be eating but so i can choose something healthy she could tell me not to eat it \\\hline
309 &	Affirm, low &	Counselor &	okay that sounds like a really great suggestion \\\hline
\end{tabular}
\end{table}

@@

\textbf {Recurrent Neural Network (RNN)} is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step. In this way, RNN can capture long-term dependencies for predicting the future events, which is the main advantage of RNN over feedforward Neural Network, MC, and HMM. The capability of remembering past event is very useful in motivational interviews where a behavior observed at some points in the interview is very informative for the future behaviors that will be observed. However, it was observed that RNN fails to capture long-term dependencies due to vanishing gradient problem \cite{bengio1993problem}. In order to mitigate this problem, Hochreiter et al.\cite{hochreiter1997long} proposed a special kind of RNN called Long Short Term Memory networks or simply LSTM. Later, the model was enhanced by including forget gates and the tanh activation function \cite{graves2013speech}. There are several variants of LSTM model, where a dramatic variation on the LSTM is the Gated Recurrent Unit\cite{cho2014properties} or simply GRU. GRUs are simpler than LSTM units and experimentally outperform than other models in many cases. We reiterate the mathematical formulation of GRU defined by Chung et al.\cite{chung2014empirical} as follows (Eq.~\ref{eq:firstgru}~to~\ref{eq:lastgru}):
\begin{equation}
z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
\label{eq:firstgru}
\end{equation}
\begin{equation}
r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
\end{equation}
\begin{equation}
\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
\end{equation}
\begin{equation}
h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
\label{eq:lastgru}
\end{equation}  
\begin{equation}
total\ loss = \alpha \cdot \frac{1}{T}\sum_{t=1}^T loss(\bar y^{(t)},y^{(t)}) + (1 - \alpha) \cdot loss(\bar y^{(T)},y^{(T)})
\label{eq:loss}
\end{equation}  
\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.90\textwidth]{figures/rnn.eps}
    \caption{\textbf{Proposed RNN model with target replication}.}
    \label{fig:rnn-model}
\end{figure}
The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}. We used one hidden LSTM or GRU layer of 20 nodes after embedding the given myscope codes from the input sequence. Then, we applied a softmax layer at each time step for predicting the target. This target replication strategy was used in several studies \cite{lipton2015learning,choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks. As it can be seen from equation~(\ref{eq:loss}), our loss is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determines the relative importance of intermediate outputs. Our model contains several hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which were determined by the validation set. We implemented our models in Tensorflow with adam optimizer and early stopping based on the validation loss. Experimentally, we observed that our model converges to optimal after 100 epoch and have no effect of dropout and regularization. The code for all models is publicly available at https://github.com/teanalab/myscope-sequential-analysis.

@@

\begin{table}[h]
\centering
\caption{Performance of MC, HMM, and RNN for predicting success of under-sampled patient-provider communication sequences. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_under_sampled}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Method} & \textbf{Accuracy}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline

@@

\subsection*{\textit{Predictive performance for under-sampled PPC code sequences}}
We applied small learning rate 0.00005 and 150 epoch with early stopping strategy for training deep learning models when small size of under-sampled data is used. Predictive performance summary of the proposed methods on under-sampled sequences is presented in Table~\ref{tab:result_under_sampled}. Five major conclusions can be drawn from the results in Table~\ref{tab:result_under_sampled}. First, recurrent neural networks outperforms over probabilistic models, and achieved 18.50\%-39.88\% higher F1-score. Second, GRU with target replication has the best performance over all other RNN methods, and achieved accuracy 0.8858, F1-score 0.8853 with precision 0.8917 and recall 0.8858. Third, target replication strategy improves the performance of GRU and LSTM, while conventional GRU shows better performance than LSTM with target replication. Fourth, among probabilistic models, MC-based method generally outperforms HMM across all metrics for under-sampled sequences. Fifth, it follows that second-order MC have lower acuracy, precision, recall and F-measure than first-order MC. In particular, acuracy, precision, recall and F-measure decrease by 15.80\%, 15.92\%, 15.80\% and 15.89\%, when going from first- to second-order models. At last, accuracy and recall illustrate exactly the same values for all models. \\

\begin{table}[h]
\centering
\caption{Performance of MC, HMM, and RNN for predicting success of over-sampled patient-provider communication sequences. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_over_sampled}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Method} & \textbf{Accuracy}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline

@@

\subsection*{\textit{Predictive performance for over-sampled PPC code sequences}}
For over-sampled data, RNN models are trained for 100 epoch with early stopping strategy. Table~\ref{tab:result_over_sampled} summarizes predictive performance of the proposed methods on over-sampled sequences. These results indicate that HMM had better performance than second-order MC, achieving 5.74\% higher F1-score, while HMM still have 2.41\% lower F1-score than first-order Markov chain. Similar to the case of under-sampled sequences, target replication improves the performance of RNN models, and GRU with target replication has the highest F1-score of 0.8716 among all models. However, different from under-sampled sequences, predictive performance of probabilistic models increases whereas recurrent neural networks performance decreases. We also observed that all models have largest value in precision compared to all other performance metrics.

>>>>

As can be seen from Table~\ref{tab:data_dist}, our data set is imbalanced and predominately composed of successful sequences with only a small percentage (17.80\%) of unsuccessful sequences. Usually, predictive accuracy misrepresents the performance of an employed algorithm for an imbalanced data. Because simply predict the majority class (successful) would provide a predictive accuracy 82.20\%. Therefore, it is important to handle imbalance data properly. Many solutions are proposed to deal with the imbalanced data sets, which can be divided into two categories: data and algorithmic levels. In this study, we utilized under and over sampling methods at data processing levels for balancing the adolescent obesity data set. Synthetic Minority Over-sampling (SMOTE) method is the most widely used method for over-sampling an imbalanced data set, in which new synthetic examples were generated from minority class \cite{chawla2002smote}. In this study, we generated synthetic examples along the borderline between minority examples and their selected nearest neighbors \cite{nguyen2011borderline}. On the other hand, under sampling method under samples the majority class by replacing a cluster of majority samples by the cluster centroid of a K-Means algorithm. Annotation column in Table~\ref{tab:anno_examp} shows the sequence of behavior codes from top to bottom, where counselor starts with an open-ended question and gets positive feedback at the end. It was noticed that same context represents different behavior codes which can be seen at row 7 and 9 in Table~\ref{tab:anno_examp}. Because human brain can remember the past behaviors, and classified the current context into different behavior codes after considering their past behaviors. In this paper, we only consider those models that can capture the past events for the classification of behavior sequences.\\

\begin{table}[h]
\caption{Fragment of the annotated transcript of a dialogue between a counselor and an adolescent.}    
\label{tab:anno_examp}
\centering
\begin{tabular}{|l|p{3.2cm}|l|p{8cm}|}
\hline
\bf{Annotation}  & \bf{Behavior} & \bf{Speaker} & \bf{Text} \\\hline
SS & Structure Session & Counselor & Okay. Can I meet with Xxxx alone for a few minutes? \\\hline
OQO & Open-ended question, other & Counselor & So, Xxxx, how you doing? \\\hline
HUPO &	High uptake, other	& Adolescent &	Fine \\\hline
OQTBN &	Open-ended question, target behavior neutral & Counselor &	That's good.  So, tell me  how do you feel about your weight? \\\hline
CHT+ &	Change Talk positive	& Adolescent &	It's not the best. \\\hline
CQECHT+ & Closed question, elicit change talk positive & Counselor & It's not the best? \\\hline
CHT+ &	Change Talk positive &	Adolescent & Yeah \\\hline
CQTBN &	Closed question, target behavior neutral  & Counselor &	Okay, so have you tried to lose weight before? \\\hline
HUPW &	High uptake, weight & Adolescent &	Yes \\\hline
\end{tabular}
\end{table}

@@

\textbf {Recurrent Neural Network (RNN)} is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step. In this way, RNN can capture long-term dependencies for predicting the future events, which is the main advantage of RNN over feedforward Neural Network, MC, and HMM. The capability of remembering past event is very useful in motivational interviews where a behavior observed at some points in the interview is very informative for the future behaviors that will be observed. However, it was observed that RNN fails to capture long-term dependencies due to vanishing gradient problem \cite{bengio1993problem}. In order to mitigate this problem, Hochreiter et al.\cite{hochreiter1997long} proposed a special kind of RNN called Long Short Term Memory networks or simply LSTM. Later, the model was enhanced by including forget gates and the tanh activation function \cite{graves2013speech}. There are several variants of LSTM model, where a dramatic variation on the LSTM is the Gated Recurrent Unit\cite{cho2014properties} or simply GRU. GRUs are simpler than LSTM units and experimentally outperform than other models in many cases. We reiterate the mathematical formulation of GRU defined by Chung et al.\cite{chung2014empirical} as follows (Eq.~\ref{eq:firstgru}~to~\ref{eq:lastgru}):
\begin{equation}
z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
\label{eq:firstgru}
\end{equation}
\begin{equation}
r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
\end{equation}
\begin{equation}
\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
\end{equation}
\begin{equation}
h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
\label{eq:lastgru}
\end{equation}  
\begin{equation}
total\ loss = \alpha \cdot \frac{1}{T}\sum_{t=1}^T loss(\bar y^{(t)},y^{(t)}) + (1 - \alpha) \cdot loss(\bar y^{(T)},y^{(T)})
\label{eq:loss}
\end{equation}  
\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.90\textwidth]{figures/rnn.eps}
    \caption{\textbf{Proposed RNN model with target replication}.}
    \label{fig:rnn-model}
\end{figure}
The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}. We used one hidden LSTM or GRU layer of 20 nodes after embedding the given myscope codes from the input sequence. Then, we applied a softmax layer at each time step for predicting the target. This target replication (TR) strategy was used in several studies \cite{lipton2015learning,choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks. As it can be seen from equation~(\ref{eq:loss}), our loss is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determines the relative importance of intermediate outputs. Our model contains several hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which were determined by the validation set. We implemented our models in Tensorflow with adam optimizer and early stopping based on the validation loss. Experimentally, we observed that our model converges to optimal after 100 epoch and have no effect of dropout and regularization. The code for all models is publicly available at https://github.com/teanalab/myscope-sequential-analysis.

@@

\begin{table}[h]
\centering
\caption{Performance of MC, HMM, and RNN for predicting the success of under-sampled patient-provider communication sequences. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_under_sampled}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Method} & \textbf{Accuracy}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline

@@

\subsection*{\textit{Predictive performance for under-sampled PPC code sequences}}
We applied small learning rate 0.00005 and early stopping strategy with 150 epoch for training deep learning models when under-sampled data is used. Predictive performance summary of the proposed methods on under-sampled sequences is presented in Table~\ref{tab:result_under_sampled}. Five major conclusions can be drawn from the results in Table~\ref{tab:result_under_sampled}. First, recurrent neural networks outperform over probabilistic models and achieved 18.50\%-39.88\% higher F1-score. Second, GRU with target replication has the best performance over all other RNN methods, and achieved accuracy 0.8858, F1-score 0.8853 with precision 0.8917 and recall 0.8858. Third, target replication strategy improves the performance of GRU and LSTM, while conventional GRU shows better performance than LSTM with target replication. Fourth, among probabilistic models, the MC-based method generally outperforms HMM across all metrics for under-sampled sequences. Fifth, it follows that second-order MC have lower accuracy, precision, recall, and F-measure than first-order MC. In particular, accuracy, precision, recall and F-measure decrease by 15.80\%, 15.92\%, 15.80\% and 15.89\%, when going from first- to second-order models. At last, accuracy and recall illustrate exactly the same values for all models. \\

\begin{table}[h]
\centering
\caption{Performance of MC, HMM, and RNN for predicting the success of over-sampled patient-provider communication sequences. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_over_sampled}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Method} & \textbf{Accuracy}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline

@@

\subsection*{\textit{Predictive performance for over-sampled PPC code sequences}}
For over-sampled data, RNN models are trained for 100 epoch with early stopping strategy. Table~\ref{tab:result_over_sampled} summarizes predictive performance of the proposed methods on over-sampled sequences. These results indicate that HMM had better performance than second-order MC, achieving 5.74\% higher F1-score, while HMM still has 2.41\% lower F1-score than first-order Markov chain. Similar to the case of under-sampled sequences, target replication improves the performance of RNN models, and GRU with target replication has the highest F1-score of 0.8716 among all models. However, the predictive performance of recurrent neural networks decreases in over-sampled sequences whereas the performance of probabilistic models decreases. We also observed that all models have the largest value in precision compared to other performance metrics.
